Mathematical Language Processing \\ Project

### 2014-03-10 paper.tex
Shrink down the paper to fit to four pages

\begin{abstract}
In this paper, we present a solution for automated definition discovery for
the variables occurring in mathematical formulae. Ordinary natural language
processing approaches that are trained based on large, manually annotated news
corpora fall short in reasonable quality of definition discovery for
scientific texts. In the \emph{Mathematical Language Processing} project, we
discover definitions
%we did not discover any relations  a relation would be something like "let
%<subj> be a <object>", or "we call <subj> <obj>", or "<subject> is <object>"
%or just "<subj> <obj>"
for identifiers in formulae from the surrounding text by statistical methods.
%with a machine learning approach for classification. This was also not done
%in this paper We extract the identifiers from the formulae and track co-
%occurrences of those in the surrounding text.
The evaluation of our prototypical system, applied on the Wikipedia text
corpus shows that our approach augments the user experience for articles with
mathematical formula substantially. While hovering the identifiers occurring
in the equation a pop-up with the most probable explanations for the
identifiers occurs. The displayed definition information provides a good match
to the actual meaning of the identifiers.
\keywords{machine learning, text mining, parallel computing}
\end{abstract}

@@

\subsection{Motivation}
Mathematical formulae are a viable source of information for a wide range of
scientists. Often those formulas contain variables that are unknown or at
least ambiguous to the reader. Therefore, one needs to study the surrounding
text to find the relevant definition.

@@

We will show how the proposed strategy can be implemented in the PACT
programming model \cite{Alexandrov2010}.

\subsection{Related Work}
\citeauthor{Quoc2010} (2010) proposed in cite{Quoc2010} an approach for
relating whole formulas to sentences and their describing paragraphs.
\citeauthor{Yokoi} (2010) trained a emph{support vector machine} to extract
natural language descriptions for mathematical expressions cite{Yokoi}.

\section{Relation Detection}
The basic assumption of our approach is that if two entities take part in a
relation, those two co-occur in the same sentence. This assumption is also
used in the \emph{Distant Supervision} \cite{Mintz2008} approach by
\citeauthor{Mintz2008}. Furthermore, we assume in a scientific document, which
contains a specific variable name, there is at least one sentence with such a
relation.

The two entities have some characteristics that can be used for filtering.
First of all, the relevant variables can be reieved accurately from \TeX\
formatted formulas (see Section \ref{vr}). The fact that a relation consists
of two entities and we know that one must be the variable, leads to an obvious
conclusion: Sentences without a variable cannot contain a definition relation
and can be discarded. Second, one can assume that the other entity must be a
noun or a compound noun. Thus, all words that are not nouns or adjectives
describing these nouns, can also be discarded.

\subsection{Numerical Statistics}
As a prerequisite for detecting potential relations, one needs to identify all
occurrences of a variable within a document. We assume that the distance
between the identifier and the corresponding definition is normal distributed.

Formally, we calculate the distance in words
$\Delta=p_\mathrm{def}-p_\mathrm{id}$  between $p_\mathrm{def}$ the position
of the definition and $p_\mathrm{id}$ the position of the identifier.

Thus, the assumed normalized relation probability density function reads
\begin{equation}
\label{eq:dist}
R_\sigma(\Delta)= {\sigma \sqrt{2\pi}}{e^{{\left(2\sigma^2\right)^{-1}}}} \mathcal N(0,\sigma^2)(\Delta)\equiv \exp\left[-\frac{1}{2}\frac{\Delta^2-1}{\sigma^2}\right].
\end{equation}
\newcommand{\normalPlot}[2]{(e^((-1/2)*(((#1)^2-1)/(#2)^2)))} %1/((#2)*sqrt(2*pi)))*
\newcommand{\normalPlotFWHM}[2]{\normalPlot{#1}{ sqrt(((#2)^2-1)/(2*ln(2)))}}

Thereby we performed the normalization, with regard to the model assumption
that the probability to find a relation at $\Delta=1$ is maximal, i.e.
$R(1)=1$.  Manual evaluation of some articles indicated that
$R(1)\approx\frac{1}{2}R(5)$. As, the full width half maximum is at
$\Delta=\sqrt{2 \ln(2)  \sigma^2+1}$, and the standard deviation evaluates to
$\sigma(\Delta)=\sqrt\frac{\Delta^2-1}{2\ln 2}$. As a consequence we choose
$\sigma_d=\sqrt\frac{5^2-1}{2\ln 2}$ and write $R_\mathrm{d}(\Delta) \equiv
R_{\sigma_\mathrm{d}}(\Delta) $. Please note that this assumption is heavily
influenced by the language of the text corpus.

Not only the distances within one sentence are of interest. Also the
sequential order of sentences, containing a variable, carry some viable
sources for statistics. Typically an author first introduces a term and uses
it in subsequent sentences without necessarily repeating its definition. We
assume that this kind of behavior also applies to the use of variables. Thus,
we extend the idea of distance to the sequential order of sentences. Here, the
distance is the count $n$ of candidate sentences, beginning at the top of the
document.

In our experiments, we expect that half of the definition relations are found
in the sentence with 3 sentences difference. Thus we choose
\begin{equation}
\label{eq:order}
R_{s}(n)={\sigma_\mathrm{s} \sqrt{2\pi}} \mathcal N(0,\sigma_\mathrm{s}^2)(n)= \exp\left[-\frac{1}{2}\frac{n^2}{\sigma_\mathrm{s}^2}\right]
\end{equation}
 with $\sigma_\mathrm{s}=\frac{3}{\sqrt{2\ln 2}}$.
\newcommand{\normalPlotFWHMII}[2]{(e^((-1/2)*(((#1)^2*2*ln(2))/(#2)^2)))}

\begin{figure}[H]
\begin{tikzpicture}
	\begin{axis}[
			axis x line = bottom,
			axis y line = left,
			height=3.5cm, width=12cm,
			xmin=-10, xmax=10,
			ymin=0, ymax=1,
			xtick={-5,-3,0,3,5},
			ytick={0,0.5,1},
			xlabel=$Distance$,
			ylabel=$Ranking$,
			domain={-21:21},
			samples at={-10,-9,...,-1,1,2,...,10}
			]
		\addplot [mark=x,smooth]{\normalPlotFWHM{x}{5}};
		\addplot [mark=x,smooth,samples at={-10,-9,...,10}]{\normalPlotFWHMII{x}{3}};
		\addplot [black,dotted] coordinates {(0,0) (0,1)};
		\addplot [black,dashed] coordinates {(5,0) (5,1)};
		\addplot [black,dashed] coordinates {(3,0) (3,1)};
		\addplot [black,dashed] coordinates {(-5,0) (-5,1)};
		\addplot [black,dashed] coordinates {(-3,0) (-3,1)};
	\end{axis}
\end{tikzpicture}
%		\addplot [mark=,domain=-20:20,smooth]{e^(-(x^2/13.5))};
%			x dir=reverse,
\caption{Curves of the rating functions $R_{d}$ and $R_{s}$}
\end{figure}

Furthermore, we take distributional properties into account. The classic tf-
idf \cite{Salton86} statistic reflects the importance of a term to a document.
Unfortunately, the \emph{inverse document frequency} is not applicable to
definition relations. The frequency distribution of all terms, that
participate in various relations is unknown. Thus, using tf-idf could affect
the sensitivity of our rating statistic significantly. E.g., `Hamiltonian' as
well as `length' are both valid entities. Obviously, the number of documents
containing one of these words, within a corpus like Wikipedia, differs a lot.

Therefore, we only use the normalized term frequency. $t$ and $w$ are terms.
$f(t,d)$ denotes the frequency of a term within a document $d$.

\begin{equation}
	\label{eq:tf}
	\mathrm{tf}(t,d) = \frac{\mathrm{f}(t,d)}{\max\{\mathrm{f}(w,d):w \in d\}}
\end{equation}

We define our ranking statistic as the normalized weighted mean of the
rankings defined in Equations \ref{eq:dist}, \ref{eq:order}, and \ref{eq:tf}:

\begin{equation}
	\label{eq:rating}
	R(n,\Delta,t,d)=\frac{\alpha{R}_{\mathrm d}(\Delta)+\beta{R}_{\mathrm s}(n)+\gamma\mathrm{tf}(t,d)}{\alpha+\beta+\gamma} \mapsto [0;1]
\end{equation}

\section{Implementation}

\subsection{Overview of the processor}
The MLP processing system \cite{github} is basically a set of PACTs, where
data flows through. In contrast to other programming models like Map/Reduce,
PACTs can operate on multiple inputs. At some point of time, one needs an
ordered list of tokenized sentences and a list of identified variables, to
compute the ranking score. A \emph{CoGroup Contract} takes both lists and
computes the subsets, build upon distinct document keys. Finally, a
\emph{Reduce Contract} filters out results below a given threshold $\theta$.

@@

\subsection{Experiments}

\subsubsection{Variable Retrieval}
\label{vr}
Throughout our experiments we made some observations that had an impact on the
accuracy of retrieving the correct set of variables. First of all, people tend
to incorrectly use \TeX\ trying to create formulas. E.g.,
\texttt{\textbackslash text\{log\}} is more often used than the correct
operator \texttt{\textbackslash log}. Another problem is that sometimes people
use indices as a form of annotation, like $T_{before}$, $R_{A}$, and $R_{B}$.
These ambiguities lead to a increased number of false positives.

@@

\subsubsection{Tuning}
The weighting parameters $\alpha$, $\beta$, and $\gamma$ in Equation
\ref{eq:rating} are used to balance the contribution of all embedded
statistics to the overall ranking. But these parameters need to be tuned
carefully. Otherwise the results tend to be biased towards one parameter. In
other words, if the distance measure $R_{D}$ dominates the other measures,
more complex\footnote{in terms of words between both entities, participating
in a relation} relations are less likely to be discovered.

The \emph{term frequency} turns out to be the least robust statistic. Usually,
the frequency of a specific term co-occurring with a variable name within one
document is rather low. Thus, the derived ranking is relatively volatile. In
our experiments, we chose $\gamma$ to be 0.75 ($\alpha = \beta = 1$) for a
more robust ranking.

@@

%460 identifiers

\bibliography{mlp-papers}
\end{document}

>>>>

\begin{abstract}
This paper presents automated definition discovery for
variables occurring in mathematical formulae. Ordinary natural language
processing approaches are trained based on large, manually annotated news
corpora. They fall short in reasonable quality of definition discovery for
scientific texts. In the \emph{Mathematical Language Processing} project, we
discover definitions
%we did not discover any relations  a relation would be something like "let
%<subj> be a <object>", or "we call <subj> <obj>", or "<subject> is <object>"
%or just "<subj> <obj>"
for identifiers in formulae from the surrounding text by statistical methods.
%with a machine learning approach for classification. This was also not done
%in this paper We extract the identifiers from the formulae and track co-
%occurrences of those in the surrounding text.
The evaluation of our prototypical system, applied on the Wikipedia text
corpus shows that our approach augments the user experience for articles with
mathematical formula substantially. While hovering the identifiers occurring
in the equation a pop-up with the most probable explanations for the
identifiers occurs. The displayed definition information provides a good match
to the actual meaning of the identifiers.
\keywords{machine learning, text mining, parallel computing}
\end{abstract}

@@

Mathematical formulae are a viable source of information for a wide range of
scientists. Often those formulas contain variables that are unknown or at
least ambiguous to the reader. Therefore, one needs to study the surrounding
text to find the relevant definition.

@@

We implement  show how the proposed strategy can be implemented in the PACT
programming model \cite{Alexandrov2010}.

\paragraph{Related Work}
\citeauthor{Quoc2010}\cite{Quoc2010} proposed an approach for
relating whole formulas to sentences and their describing paragraphs.
\citeauthor{Yokoi}\cite{Yokoi} trained a emph{support vector machine} to extract
natural language descriptions for mathematical expressions.

\section{Statistical definition discovery}
We detect relations between identifiers and their description in two steps.
First, we extract the identifiers from the display-style formulae and
second we determine their description from the surrounding text.

Inspired by the \emph{Distant Supervision} approach \cite{Mintz2008},
we assume that the identifier and the definition co-occur in the same sentence.
Furthermore, we restrict ourselves to identifiers that co-occur with a noun
phrase in one sentence in the document at least.

For each description candidate we use the weighted sum
\begin{equation} \label{eq:rating}
	R(n,\Delta,t,d)=\frac{\alpha{R}_{\sigma_\mathrm d}(\Delta)
		+\beta{R}_{\sigma_\mathrm s}(n)
		+\gamma\mathrm{tf}(t,d)}{\alpha+\beta+\gamma} \mapsto [0;1]
\end{equation}
that respects the distance $\Delta$ between identifier and definition term $t$, the distance $n$ between the display-style equation and th sentence that contains the term and the identifier and the term frequency $\mathrm{tf}(t,d)$ in the current document $d$ 
with a normal distributed distance measure $R_\sigma(\Delta)= \exp\left[-\frac{1}{2}\frac{\Delta^2-1}{\sigma^2}\right].$
Our model assumption is that the probability to find a relation at $\Delta=1$ like for example in \emph{the energy $E$ and the mass $m$} is maximal, i.e. $R_{\sigma_\mathrm d}(1)=1$.  In order to determine the full width half maximum of our distribution we evaluated some articles manually and found $R_{\sigma_\mathrm d}(1)\approx 2 R_{\sigma_\mathrm d}(5)$ and thus $\sigma_d=\sqrt\frac{5^2-1}{2\ln 2}$.

The probability to find a correct definition decays for 50 \% within three sentences. Consequently  $\sigma_\mathrm s=2\left({\ln 2}\right)^{-\frac{1}{2}}$.

The classic tf-idf \cite{Salton86} statistic reflects the importance of a term to a document. For our task the inverse document frequency (idf)  would assign high penalties to frequent which are valid definitions for identifiers like in the example `Energy` and `Einstein`. Consequently we used just the term frequency which still had an impact on the overall result that seemed to high in our manual experiments. Therefore, we use the tuning parameters $\gamma=0.75$ and $\alpha = \beta = 1$.

%The weighting parameters $\alpha$, $\beta$, and $\gamma$ in Equation
%\ref{eq:rating} are used to balance the contribution of all embedded
%statistics to the overall ranking. But these parameters need to be tuned
%carefully. Otherwise the results tend to be biased towards one parameter. In
%other words, if the distance measure $R_{D}$ dominates the other measures,
%more complex\footnote{in terms of words between both entities, participating
%in a relation} relations are less likely to be discovered.
%
%The \emph{term frequency} turns out to be the least robust statistic. Usually,
%the frequency of a specific term co-occurring with a variable name within one
%document is rather low. Thus, the derived ranking is relatively volatile. In
%our experiments, we chose $\gamma$ to be 0.75 ($\alpha = \beta = 1$) for a
%more robust ranking.
%to  E.g., `Hamiltonian' as well as `length' are both valid entities. Obviously, the number of documents containing one of these words, within a corpus like Wikipedia, differs a lot.
%
%Therefore, we only use the normalized term frequency. $t$ and $w$ are terms.
%$f(t,d)$ denotes the frequency of a term within a document $d$.
%
%\begin{equation}
%	\label{eq:tf}
%	\mathrm{tf}(t,d) = \frac{\mathrm{f}(t,d)}{\max\{\mathrm{f}(w,d):w \in d\}}
%\end{equation}
%

\section{Implementation}
The MLP processing system \cite{github} is basically a set of PACTs, where
data flows through. In contrast to other programming models like Map/Reduce,
PACTs can operate on multiple inputs. At some point of time, one needs an
ordered list of tokenized sentences and a list of identified variables, to
compute the ranking score. A \emph{CoGroup Contract} takes both lists and
computes the subsets, build upon distinct document keys. Finally, a
\emph{Reduce Contract} filters out results below a given threshold $\theta$.

@@

%\subsection{Experiments}
\section{Evaluation}
%\subsubsection{Variable Retrieval}
\label{vr}
Throughout our experiments we made some observations that had an impact on the
accuracy of retrieving the correct set of variables. First of all, people tend
to incorrectly use \TeX\ trying to create formulas. E.g.,
\texttt{\textbackslash text\{log\}} is more often used than the correct
operator \texttt{\textbackslash log}. Another problem is that sometimes people
use indices as a form of annotation, like $T_{before}$, $R_{A}$, and $R_{B}$.
These ambiguities lead to a increased number of false positives.

@@

@@

%460 identifiers
\begingroup
\let\clearpage\relax
\bibliography{mlp-papers}
\endgroup

\end{document}
